Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-05-25 Paper/paper.tex
intro first pass

\red{
BEgin this paragraph by pointing out that EOS is universal among all NSs. 
Therefore, one universal $\lambdans$ vs $\mns$ curve. Therefore, by combining
multiple BHNS events, expect to measure $\lambdans$ better than with any 
single BHNS event.
% \prayush{
Multiple realistic observations deliver far more information than what a single
observation furnishes, as has been shown the case to be for binary neutron
stars~\cite{DelPozzo:13}. As our final study, we perform a fully-Bayesian 
analysis of how our estimation of $\lambdans$ changes as we accumulate more and
more detections.
\red{
MOVE into later sections. Keep here only 1-2 informational sentences.
\prayush{
The choices that determine the population are important here,
and were made as follows. Neutron star mass is fixed for signals at
$1.35M_\odot$, and its spin $\chins=0$. Black hole mass is sampled uniformly
from the range $[2,5]\times 1.35=[2.7, 6.75]M_\odot$. Our choice here is given
by the intersection set of the mass range that allows for neutron star disruption
and the range supported by the waveform model LEA+~\cite{Foucart2012,
Foucart:2013a,Lackey:2013axa}. Black hole spin is sampled uniformly from the
{\it aligned} range, i.e. $\chibh\in[0,+1]$. The true deformability is fixed
once per population, sampled uniformly from the range $\lambdans\in[0,2000]$.
In order to keep the computational cost tractable, we make a further 
approximation.
\textbf{%
We inject signals into zero noise at different SNRs and recover
their parameters from Bayesian inferencing for a fixed set of binaries,
whose parameters lie on a uniform grid given by:
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
Once the population parameters been selected, each event's parameters are
replaced by their nearest neighbor in the above set.
We make sure that this approximation is conservative by keeping our population
sampling ranges within the conservative side of the range spanned by the above
set. With the caveats laid out, we report our results in
Sec.~\ref{s1:multiple_observations}.
\red{
MOVE to SUMMARY, only keeping 2-4 sentences here. Also, paradigm A/B barely
matters. Therefore, first describe the common results, then point out how the
presence/absence of mass gap modifies them.
\prayush{
In addition, we probe two paradigms, one
that allows for BH masses to lie within the astrophysical mass-gap (paradigm
A), and one that does not (paradigm B).
For paradigm A, we find the following: (i) for the softer equations of state
that result in less deformable neutron stars, $15-20$ detections are sufficient
for our median measured value of $\lambdans$ to lie within $10\%$ of the true
value. (ii) For NSBH populations with more deformable NSs
($\lambdans\geq 1000$), the same is achievable within as few as $10$ realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans\geq 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs do indeed have
masses between $\leq 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
For paradigm B, we find the information accumulation to be somewhat slower.
While the quantitative inferences for $\lambdans\geq1000$ populations are
not affected significantly, we find $\mathcal{O}(10-20)\%$ more events are
required to attain the same measurement accuracy as under paradigm A.
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, with a few years'
time-scale.

Finally, we note that our results are applicable for the design sensitivity
LIGO instruments, that are expected to come online in $2018-19$.
\red{
MOVE towards end of SUMMARY, into ``Future work''.
}
\prayush{
Also note
that more a recent work that improves upon the waveform model we use in this
study has been published~\cite{Pannarale:2015jka}, but it provides only an 
amplitude model which needs to be (in future) augmented with a compatible
phase model. In addition, accuracy of the underlying numerical simulations 
used to calibrate the waveform model used here have not been verified against 
independent codes so far.
It is therefore difficult to assess the combined modeling error and its effect
on our results. they are, therefore, limited by the limitations of our
waveform model. However, we expect the combined effect of modeling errors to
not change our qualitative conclusions.
}

\red{
The following numerical details disrupt the flow of the text. MOVE after
description of analyses.
}
\prayush{
As in rest of the paper, we restrict here our parameter space to span 
mass-ratios $q:=\mbh/\mns\in[2,5]$, black hole spin (aligned with orbit)
$\chibh\in[-0.5, +0.75]$, and dimensionless neutron star tidal deformability 
$\lambdans:= G\left(\frac{c^2}{G \mns}\right)^5\lambda \in[500, 2000]$.
The mass of NS for signals (and not templates) is fixed to $1.35M_\odot$ 
throughout~\cite{stellarcollapsemass}, and is taken to be 
non-spinning~\cite{Miller:2014aaa}.
Each signal is rescaled to a range of SNR values $\rho\in[10, 70]$.
}

@@

>>>>

% \red{
% BEgin this paragraph by pointing out that EOS is universal among all NSs. 
% Therefore, one universal $\lambdans$ vs $\mns$ curve. Therefore, by combining
% multiple BHNS events, expect to measure $\lambdans$ better than with any 
% single BHNS event.
% }
% % \prayush{
The EoS of matter at nuclear densities is believed to be universal among all
neutron stars. The TOV equations therefore furnish a unique dependence of
$\lambdans$ on $\mns$. This fact enables us to go further - combine
information from multiple observations of realistic NSBH systems and
perform a fully-Bayesian analysis of how our estimation of $\lambdans$ changes
as we accumulate detections. This is similar to an earlier
study~\cite{DelPozzo:13} aimed at binary neutron stars.
% % }
% % 
% \red{
% MOVE into later sections. Keep here only 1-2 informational sentences.
% }
% % 
We restrict ourselves to a population of NSs with masses clustered very tightly
around $1.35M_\odot$ (with a negligible variance), and negligible spins. We
sample different nuclear EoSs by sampling entire populations with different
values set for the tidal deformability of the NS.
% These values correspond to a hard EoS at the lower end, and a soft EoS at the upper.
Finally, we take source
locations to be uniformly distributed in spatial volume, and source
orientations to be uniform on a $2-$sphere. From an ensemble of such
populations, we learn the following.
% 
Our best estimate for $\lambdans$ improves much faster than our statistical
uncertainties. Within $10-20$ detections, the median measured value of
$\lambdans$ lies within $10\%$ of its actual value.
Measurement uncertainties for $\lambdans$, on the other hand, depend on
$\lambdans$ itself. We find that for hard equations of state (with 
$\lambdans\geq 1000$), $10-20$ observations are sufficient to constrain
$\lambdans$ within $\pm 50\%$. For softer equations of state, the same level
of certainty would require substantially more ($25-40$) observations.
% 
Further, if the astrophysical ``mass-gap'' is real, we find that $20-50\%$
additional observations would be required to attain the same measurement
accuracy as above. Putter tighter constraints on the $\lambdans$ of a
population would require $50+$ NSBH observations, in either scenario.
% 
All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}.

% \red{
% MOVE to SUMMARY, only keeping 2-4 sentences here. Also, paradigm A/B barely
% matters. Therefore, first describe the common results, then point out how the
% presence/absence of mass gap modifies them.
% }
% 
% % \prayush{
% In addition, we probe two paradigms, one
% that allows for BH masses to lie within the astrophysical mass-gap (paradigm
% A), and one that does not (paradigm B).
% % 
% For paradigm A, we find the following: (i) for the softer equations of state
% that result in less deformable neutron stars, $15-20$ detections are sufficient
% for our median measured value of $\lambdans$ to lie within $10\%$ of the true
% value. (ii) For NSBH populations with more deformable NSs
% ($\lambdans\geq 1000$), the same is achievable within as few as $10$ realistic
% observations. (iii) The statistical uncertainty associated with $\lambdans$
% measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
% when $\lambdans\geq 1000$), and using $25-40$ observations for softer equations
% of state. All of the above is possible within a few years of design
% aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs do indeed have
% masses between $\leq 5M_\odot$ (i.e. in the mass-gap). However, further
% restricting $\lambdans$ will require $50+$ NSBH observations.
% % 
% For paradigm B, we find the information accumulation to be somewhat slower.
% While the quantitative inferences for $\lambdans\geq1000$ populations are
% not affected significantly, we find $\mathcal{O}(10-20)\%$ more events are
% required to attain the same measurement accuracy as under paradigm A.
% % 
% We conclude that within as few as $20-30$ observations of disruptive NSBH
% mergers, aLIGO will begin to place interesting bounds on NS deformability.
% This, amongst other things, will allow us to rank different equations of 
% state for neutron star matter from most to least likely, with a few years'
% time-scale.
% % }

% \red{
% The following numerical details disrupt the flow of the text. MOVE after
% description of analyses.
% }

% \prayush{
In rest of the paper, we restrict our parameter space to span 
mass-ratios $q:=\mbh/\mns\in[2,5]$, black hole spin (aligned with orbit)
$\chibh\in[-0.5, +0.75]$, and dimensionless neutron star tidal deformability 
$\lambdans:= G\left(\frac{c^2}{G \mns}\right)^5\lambda \in[500, 2000]$.
The mass of NS for signals (and not templates) is fixed to $1.35M_\odot$ 
throughout~\cite{stellarcollapsemass}, and is taken to be 
non-spinning~\cite{Miller:2014aaa}.
Each signal is rescaled to a range of SNR values $\rho\in[10, 70]$.
% }
% 
Finally, we note that our results are applicable for the design sensitivity
LIGO instruments, that are expected to come online in $2018-19$.

@@

\prayush{
The choices that determine the population are important here,
and were made as follows. Neutron star mass is fixed for signals at
$1.35M_\odot$, and its spin $\chins=0$. Black hole mass is sampled uniformly
from the range $[2,5]\times 1.35=[2.7, 6.75]M_\odot$. Our choice here is given
by the intersection set of the mass range that allows for neutron star disruption
and the range supported by the waveform model LEA+~\cite{Foucart2012,
Foucart:2013a,Lackey:2013axa}. Black hole spin is sampled uniformly from the
{\it aligned} range, i.e. $\chibh\in[0,+1]$. The true deformability is fixed
once per population, sampled uniformly from the range $\lambdans\in[0,2000]$.
In order to keep the computational cost tractable, we make a further 
approximation.
%
\textbf{%
We inject signals into zero noise at different SNRs and recover
their parameters from Bayesian inferencing for a fixed set of binaries,
whose parameters lie on a uniform grid given by:
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
% 
Once the population parameters been selected, each event's parameters are
replaced by their nearest neighbor in the above set.
}
We make sure that this approximation is conservative by keeping our population
sampling ranges within the conservative side of the range spanned by the above
set. With the caveats laid out, we report our results in
Sec.~\ref{s1:multiple_observations}.
}

\prayush{
In addition, we probe two paradigms, one
that allows for BH masses to lie within the astrophysical mass-gap (paradigm
A), and one that does not (paradigm B).
% 
For paradigm A, we find the following: (i) for the softer equations of state
that result in less deformable neutron stars, $15-20$ detections are sufficient
for our median measured value of $\lambdans$ to lie within $10\%$ of the true
value. (ii) For NSBH populations with more deformable NSs
($\lambdans\geq 1000$), the same is achievable within as few as $10$ realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans\geq 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs do indeed have
masses between $\leq 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
% 
For paradigm B, we find the information accumulation to be somewhat slower.
While the quantitative inferences for $\lambdans\geq1000$ populations are
not affected significantly, we find $\mathcal{O}(10-20)\%$ more events are
required to attain the same measurement accuracy as under paradigm A.
% 
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, with a few years'
time-scale.
}

\red{
MOVE towards end of SUMMARY, into ``Future work''.
}
\prayush{
Also note
that more a recent work that improves upon the waveform model we use in this
study has been published~\cite{Pannarale:2015jka}, but it provides only an 
amplitude model which needs to be (in future) augmented with a compatible
phase model. In addition, accuracy of the underlying numerical simulations 
used to calibrate the waveform model used here have not been verified against 
independent codes so far.
It is therefore difficult to assess the combined modeling error and its effect
on our results. they are, therefore, limited by the limitations of our
waveform model. However, we expect the combined effect of modeling errors to
not change our qualitative conclusions.
}
